### 2022-11-15 NSF2022/Description/specificaim2.tex

To briefly describe IE formulations of PDEs, we consider the Stokes
equation~\eqref{eqn:stokes} as an example. Its solution can be written
as a convolution 
\begin{align}
  \uu(\xx) = \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
  \label{eqn:conv}
\end{align}
where $G(\xx - \yy)$ is a carefully chosen derivative of the PDE's
fundamental solution and $\ssigma: \partial \Omega(t) \rightarrow \RR^2$
is an unknown density function. The integral
representation~\eqref{eqn:conv} is known as a layer potential.
%This is analogous to the problem of
%finding the global electrostatic potential $u$ with surface charge
%density $\sigma$
%\begin{align*}
%  u(x) = -\frac{1}{4\pi}\int_{\partial \Omega(t)} 
%    \frac{1}{|\xx - \yy|}\sigma(y) \, ds_\yy.
%\end{align*}
The IE reformulation has several immediate advantages including a
dimension reduction since $\ssigma$ is defined only on $\partial
\Omega(t)$, and the far field boundary condition of the PDE is
automatically satisfied. The density function $\ssigma$ is chosen to
satisfy the boundary condition, and this results in an IE 
\begin{align}
  \gg(\xx) = \lambda \ssigma(\xx) + 
    \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, ds_\yy,
    \quad \xx \in \partial\Omega(t),
  \label{eqn:bie}
\end{align} where $\gg$ is the boundary condition, and $\lambda$ is a
constant that depends on the choice of $G$. For numerical stability,
choices of $G$ that result in second-kind Fredholm integral equations,
meaning $\lambda \neq 0$, are preferred.

As an illustration, the problem of finding the global electrostatic
potential $u$ with surface charge density $\sigma$ is a special case of
\eqref{eqn:conv}, where $G(\xx - \yy)$  takes the form
$-(2\pi \epsilon_0)^{-1}\log|\xx - \yy|$
in two dimensions and 
$(4\pi \epsilon_0)^{-1}|\xx -\yy|^{-1}$ in three dimensions
with $\epsilon_0$ the permittivity of vacuum.  In
this context, equation~\eqref{eqn:bie} is analogous to Gauss's Law.

@@

\begin{wrapfigure}[16]{r}{2.1in}
  \vspace{-4pt}
  \centering
  \input{figures/near_sing.tikz}
  \vspace{-15pt}
  \caption{\label{fig:interpquad} \footnotesize An interpolation-based
  quadrature method that works in two and three dimensions. The
  spectrally-accurate trapezoid rule is used in points in
  $\mathtt{Far}(\Gamma_i)$, while one dimensional interpolant is used
  for points in $\mathtt{Near}(\Gamma_i)$. $h$ is the arclength
  spacing.}
\end{wrapfigure}
Since we are interested in round (2D) and spherical (3D) granules, we
will exploit the fact that integrals can be computed with spectral
accuracy by transforming the integrals to to a Fourier (2D) or spherical
harmonic (3D) basis. This technique can be used to evaluate layer
potentials for both Stokes~\cite{cor-vee2018} and the screened Laplace
equation~\cite{koh-cor-che-vee2021}. However, this quadrature is not
appropriate or other shapes, and an alternative approach is necessary.
The Barycentric-type formulae for Cauchy integrals results in a
uniformly bounded quadrature error, but the method is described only for
the two-dimensional Laplace and Stokes equations. PI Quaife recently
showed that these quadrature methods can be extended to other
two-dimensional PDEs~\cite{che-lin-her-qua2022}. The method uses
singularity subtraction to write the fundamental solution of the PDE as a
sum of two terms: a logarithmic function and a continuously
differentiable function. Then the Barycentric-type formulae can be
applied to the logarithmic function, and the standard trapezoid
rule~\cite{tre-wei2014} can be applied to the other function. These
singularity subtraction methods can also accurately compute derivatives
of layer potentials~\cite{car2021, car2020, car-kha-kim2018} which is
required, for example, when computing the phase
stress~\eqref{eqn:stressbalance}. Singularity subtraction methods can
also be applied in three dimensions, but not combined with
Barycentric-type formulae. Alternatively, smoothed variations of the
fundamental solution can be used to address nearly-singular
integrals~\cite{beale1, cor2001, tlu-bea2019}. Finally, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}.

@@

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional required
computational tool is a collision-aware adaptive time stepping method.
The phase field results in the granules attracting one another and
this leads to nearly contacting boundaries. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when 
\begin{wrapfigure}[26]{l}{1.8in}
  \vspace{-0pt}
  \centering
  \input{figures/doublewell.tikz} 
  \vspace{-10pt}
  \caption{\label{fig:CA} \footnotesize (a) A double-well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}
simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2021}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.

@@

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 5.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (F'(\phi^N))^{-1} F(\phi^N)$, where
$F'(\phi^N) = -\epsilon^2 \Delta + W''(\phi^N)$ is the
Fr\'{e}chet derivative of $F$, and $\alpha$ is the step size that will
be determined using backtracking. This requires solving the variable
coefficient PDE $-\epsilon^2 \Delta v + W''(\phi^{N}) v =
F[\phi^N]$, which is challenging with an integral equation since a
fundamental solution is not readily available. However, by choosing a
double-well potential with $W''(\phi_0) = W''(\phi_1) = \beta$, we
propose to instead use an inexact Newton's method where $(F')^{-1}$ is
approximated with the solution of $-\epsilon^2 \Delta v +
\beta v = F[\phi^N]$. We test the efficacy of the exact and
inexact Newton's method applied to~\eqref{eqn:phaseRoot} in $(0,1)
\subset \RR$ by discretizing $F'$ with a second-order finite difference
stencil. The double-well potential is illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

>>>>

To briefly describe IE formulations of PDEs, we consider the Stokes
equation~\eqref{eqn:stokes} as an example. Its solution can be written
as a convolution 
\begin{align}
  \uu(\xx) = \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, dS_\yy,
  \label{eqn:conv}
\end{align}
where $G(\xx - \yy)$ is a carefully chosen derivative of the PDE's
fundamental solution and $\ssigma: \partial \Omega(t) \rightarrow \RR^2$
is an unknown density function. The integral
representation~\eqref{eqn:conv} is known as a layer potential.
%This is analogous to the problem of
%finding the global electrostatic potential $u$ with surface charge
%density $\sigma$
%\begin{align*}
%  u(x) = -\frac{1}{4\pi}\int_{\partial \Omega(t)} 
%    \frac{1}{|\xx - \yy|}\sigma(y) \, ds_\yy.
%\end{align*}
The IE reformulation has several immediate advantages including a
dimension reduction since $\ssigma$ is defined only on $\partial
\Omega(t)$, and the far field boundary condition of the PDE is
automatically satisfied. The density function $\ssigma$ is chosen to
satisfy the boundary condition, and this results is an IE 
\begin{align}
  \gg(\xx) = \lambda \ssigma(\xx) + 
    \int_{\partial\Omega(t)} G(\xx - \yy) \ssigma(\yy) \, dS_\yy,
    \quad \xx \in \partial\Omega(t),
  \label{eqn:bie}
\end{align} where $\gg$ is the boundary condition, and $\lambda$ is a
constant that depends on the choice of $G$. For numerical stability,
choices of $G$ that result in second-kind Fredholm integral equations,
meaning $\lambda \neq 0$, are preferred.

As an illustration, the problem of finding the global electrostatic
potential $u$ with surface charge density $\sigma$ is a special case
of~\eqref{eqn:conv}, where $G(\xx - \yy)$ takes the form $-(2\pi
\epsilon_0)^{-1}\log|\xx - \yy|$ in two dimensions and $(4\pi
\epsilon_0)^{-1}|\xx -\yy|^{-1}$ in three dimensions with $\epsilon_0$
the permittivity in a vacuum. In this context, equation~\eqref{eqn:bie}
is analogous to Gauss's Law.

@@

\begin{wrapfigure}[17]{r}{1.9in}
  \vspace{-4pt}
  \centering
  \input{figures/near_sing.tikz}
  \vspace{-15pt}
  \caption{\label{fig:interpquad} \footnotesize An interpolation-based
  quadrature method that works in two and three dimensions. The accurate
  trapezoid rule is used in points in $\mathtt{Far}(\Gamma_i)$, while
  one dimensional interpolant is used for points in
  $\mathtt{Near}(\Gamma_i)$. $h$ is the arclength spacing.}
\end{wrapfigure}
Since we are interested in round (2D) and spherical (3D) granules, we
will exploit the fact that integrals can be computed with spectral
accuracy by transforming the integrals to a Fourier (2D) or spherical
harmonic (3D) basis. This technique can be used to evaluate layer
potentials for both Stokes~\cite{cor-vee2018} and the screened Laplace
equation~\cite{koh-cor-che-vee2021}. However, this quadrature is not
appropriate or other shapes, and an alternative approach is necessary.
The Barycentric-type formulae for Cauchy integrals results in a
uniformly bounded quadrature error, but the method is described only for
the two-dimensional Laplace and Stokes equations. PI Quaife recently
showed that these quadrature methods can be extended to other
two-dimensional PDEs~\cite{che-lin-her-qua2022}. The method uses
singularity subtraction to write the fundamental solution of the PDE as
a sum of two terms: a logarithmic function and a continuously
differentiable remainder. Then the Barycentric-type formulae can be
applied to the logarithmic function, and the standard trapezoid
rule~\cite{tre-wei2014} can be applied to the remainder. These
singularity subtraction methods can also accurately compute derivatives
of layer potentials~\cite{car2021, car2020, car-kha-kim2018} which is
required, for example, when computing the phase
stress~\eqref{eqn:stressbalance}. Singularity subtraction methods can
also be applied in three dimensions, but not combined with
Barycentric-type formulae. Alternatively, smoothed variations of the
fundamental solution can be used to address nearly-singular
integrals~\cite{beale1, cor2001, tlu-bea2019}. Finally, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}.

@@

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional required
computational tool is a collision-aware adaptive time stepping method.
The phase field results in the granules attracting one another and this
leads to nearly contacting boundaries that, even with high-order spatial
accuracy, can collide due to time stepping error. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating
\begin{wrapfigure}[26]{l}{1.8in}
  \vspace{-0pt}
  \centering
  \input{figures/doublewell.tikz} 
  \vspace{-10pt}
  \caption{\label{fig:CA} \footnotesize (a) A double-well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}
dense suspensions. Therefore, we will develop a high-order
collision-avoiding adaptive time stepping method. This method will avoid
collisions by using the minimal separation distance to impose a
non-overlapping constraint. This results in a linear complementarity
problem that can be solve efficiently with a projected gradient descent
method~\cite{Yan2019}. High-order time adaptivity will be based on PI
Quaife's previous work on suspensions~\cite{qua-bir2016}, and this
method has proven to be a robust method to simulate dense
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017, qua-gan-you2021}.
Following this previous work, the proposed work will use a spectral
deferred correction method that iteratively applies a low-order,
single-step method to achieve high-order accuracy. At each time step, we
will compute the total force and torque of the system which is
computationally cheap and physically zero, and therefore provides an
estimate of the time stepping error. PI Quaife's experience is that
error estimates based on physical constraints, such as force- and
torque-free conditions, appropriately determine adaptive time step sizes
so that the dynamics are resolved without the computational expense of
techniques such as embedded Runge-Kutta, step-doubling, and Richardson
extrapolation.

@@

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 5.} How can IEs be coupled with a Newton method to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (F'(\phi^N))^{-1} F(\phi^N)$, where
$F'(\phi^N) = -\epsilon^2 \Delta + W''(\phi^N)$ is the
Fr\'{e}chet derivative of $F$, and $\alpha$ is the step size that will
be determined using backtracking. This requires solving the variable
coefficient PDE $-\epsilon^2 \Delta v + W''(\phi^{N}) v =
F[\phi^N]$, which is challenging with an integral equation since a
fundamental solution is not readily available. However, by choosing a
double-well potential with $W''(\phi_0) = W''(\phi_1) = \beta$, we
propose to instead use an inexact Newton's method where $(F')^{-1}$ is
approximated with the solution of $-\epsilon^2 \Delta v +
\beta v = F[\phi^N]$. We test the efficacy of the exact and
inexact Newton's method applied to~\eqref{eqn:phaseRoot} in $(0,1)
\subset \RR$ by discretizing $F'$ with a second-order finite difference
stencil. The double-well potential is illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.
